### 2021-07-21 paper/localpsf.tex
edits to rational positive definite section

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) A(y,x) u(x) dx dy,
\end{equation}
with integral kernel $A:\Omega \times \Omega \rightarrow \mathbb{R}$. Here $\mathcal{A}u \in L^2(\Omega)'$ is the linear functional that results from applying $\mathcal{A}$ to $u\in L^2(\Omega)$, and $\left(\mathcal{A}u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $A(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary functions (i.e., operator actions). That is, evaluation of the maps
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	A(y,x) \ge 0.
\end{equation*}
\item[Local support:] Let $\phi_x$ be the following \emph{impulse response}:
\begin{equation}
\label{eq:impulse_response_defn}
\phi_x(y) := A(y, x).
\end{equation} 
We say that $\mathcal{A}$ has local support if, for all $x\in \Omega$, the support of $\phi_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	A(y+h, x+h) \approx A(y,x).
\end{equation}
\end{description}

@@

\begin{thm}[Optimal interpolation]
	\label{thm:smoothest_interpolant}
Let $y \in \mathbb{R}^d$, and let $\gamma_y$ be the function
\begin{equation}
\label{eq:u_defn_in_thm}
\gamma_y(x) := \sum_{i=1}^r \varphi_i(y)w_i(x) = \widetilde{A}\left(y+x,x\right),
\end{equation}
where the weighting functions $w_i$ are the solutions to optimization problem \eqref{eq:wi_optimization_problem}. Then $\gamma_y$ solves the optimization problem
\begin{equation}
\label{eq:uy_optimization_problem}
\begin{aligned}
\min_{\gamma_y \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta \gamma_y}_{L^2(\Omega)}^2 \\
\text{such that} &\quad \gamma_y(x_i) = \varphi_i(y), \quad i=1,\dots,q.
\end{aligned}
\end{equation}
\end{thm}

@@

We now describe how to efficiently evaluate $\widetilde{A}(y,x)$ at an arbitrary point $(x,y) \in \Omega \times \Omega$. From \eqref{eq:kernel_entries}, we have
\begin{equation}
\label{eq:eval_y_x}
\widetilde{A}(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x),
\end{equation}
and from \eqref{eq:varphi_eval}, we have
\begin{equation}
\label{eq:kernel_entry_formua}
\varphi_{x_i}(y-x) = \begin{cases}
\eta^{b(i)}(y - x + x_i), & y - x + x_i \in E_{x_i} \\
0, & \text{otherwise},
\end{cases}
\end{equation}
where $b(i)$ denotes the batch index such that $x_i \in S^{b(i)}$. To compute $z=\widetilde{A}(y,x)$, we start with $z=0$, then loop through all batches $b$, and all sample points in each batch. For each sample point, we add one term in the sum in \eqref{eq:eval_y_x} to $z$, using \eqref{eq:kernel_entry_formua} as needed to evaluate $\varphi_{x_i}$. This is shown in Algorithm \ref{alg:eval_y_x}.

@@

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the spectrum of $\widetilde{\mathbf{A}}_\text{sym}^+$ has certain desirable properties, which we will describe later in this section. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should typically be avoided, here it is fine to compute the inverse because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix, and there are fast, stable, scalable, and widely used algorithms for inverting H-matrices \cite{Hmatrixinverse}.

@@

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $f(0)=0$ is the unique minimum of $f$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $f$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $f(\widetilde{\mathbf{A}}_\text{sym}) = \widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. The equation $f(0)=0$ implies that the null space of $\widetilde{\mathbf{A}}_\text{sym}^+$ contains the null-space of $\widetilde{\mathbf{A}}_\text{sym}$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. The equation $f(\lambda_\text{max})=\lambda_\text{max}$ forces the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ to be close to the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$. I.e., $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. The equation $f(\lambda_\text{min})=|\lambda_\text{min}|$ ensures that the negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(\lambda_\text{min})$, then the positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ will be closer to the corresponding positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ (good), but the erroneous negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ will be amplified more in magnitude (bad). We find that $f(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

>>>>

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain in dimension $d=1$, $2$, or $3$, and let $\mathcal{A}:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the form
\begin{equation}
\label{eq:kernel_representation}
(\mathcal{A}u)(v) := \int_\Omega \int_\Omega v(y) \Phi(y,x) u(x) dx dy,
\end{equation}
with integral kernel $\Phi:\Omega \times \Omega \rightarrow \mathbb{R}$. Here $\mathcal{A}u \in L^2(\Omega)'$ is the linear functional that results from applying $\mathcal{A}$ to $u\in L^2(\Omega)$, and $\left(\mathcal{A}u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We focus on operators of this form which have the following properties:
\begin{description}
\item[Matrix-free:] One cannot easily access kernel entries $\Phi(y,x)$. Access to $\mathcal{A}$ is available only through application of $\mathcal{A}$ and $\mathcal{A}^T$ to arbitrary functions (i.e., operator actions). That is, evaluation of the maps
\begin{equation*}
	u \mapsto\mathcal{A}u \quad \text{and} \quad v \mapsto\mathcal{A}^Tv.
\end{equation*} 
\item[Non-negative kernel:] For all $(y,x) \in \Omega \times \Omega$, we have
\begin{equation*}
	\Phi(y,x) \ge 0.
\end{equation*}
\item[Local support:] Let $\phi_x$ be the following \emph{impulse response}:
\begin{equation}
\label{eq:impulse_response_defn}
\phi_x(y) := \Phi(y, x).
\end{equation} 
We say that $\mathcal{A}$ has local support if, for all $x\in \Omega$, the support of $\phi_x$ is contained (or approximately contained) in a neighborhood of $x$. The smaller these supports are, the better our algorithm will perform.
\item[Local translation invariance:] If $h$ is not too large, then
\begin{equation}
	\label{eq:local_translation_invariance}
	\Phi(y+h, x+h) \approx \Phi(y,x).
\end{equation}
\end{description}

@@

\begin{thm}[Optimal interpolation]
	\label{thm:smoothest_interpolant}
Let $y \in \mathbb{R}^d$, and let $\gamma_y$ be the function
\begin{equation}
\label{eq:u_defn_in_thm}
\gamma_y(x) := \sum_{i=1}^r \varphi_i(y)w_i(x) = \widetilde{\Phi}\left(y+x,x\right),
\end{equation}
where the weighting functions $w_i$ are the solutions to optimization problem \eqref{eq:wi_optimization_problem}. Then $\gamma_y$ solves the optimization problem
\begin{equation}
\label{eq:uy_optimization_problem}
\begin{aligned}
\min_{\gamma_y \in H^2_N(\Omega)} &\quad \frac{1}{2}\norm{\Delta \gamma_y}_{L^2(\Omega)}^2 \\
\text{such that} &\quad \gamma_y(x_i) = \varphi_i(y), \quad i=1,\dots,q.
\end{aligned}
\end{equation}
\end{thm}

@@

We now describe how to efficiently evaluate $\widetilde{\Phi}(y,x)$ at an arbitrary point $(x,y) \in \Omega \times \Omega$. From \eqref{eq:kernel_entries}, we have
\begin{equation}
\label{eq:eval_y_x}
\widetilde{\Phi}(y,x) = \sum_{i=1}^r w_i(x) \varphi_{x_i}(y-x),
\end{equation}
and from \eqref{eq:varphi_eval}, we have
\begin{equation}
\label{eq:kernel_entry_formua}
\varphi_{x_i}(y-x) = \begin{cases}
\eta^{b(i)}(y - x + x_i), & y - x + x_i \in E_{x_i} \\
0, & \text{otherwise},
\end{cases}
\end{equation}
where $b(i)$ denotes the batch index such that $x_i \in S^{b(i)}$. To compute $z=\widetilde{\Phi}(y,x)$, we start with $z=0$, then loop through all batches $b$, and all sample points in each batch. For each sample point, we add one term in the sum in \eqref{eq:eval_y_x} to $z$, using \eqref{eq:kernel_entry_formua} as needed to evaluate $\varphi_{x_i}$. This is shown in Algorithm \ref{alg:eval_y_x}.

@@

First, we use fast H-matrix addition to symmetrize $\widetilde{\mathbf{A}}$:
$$\widetilde{\mathbf{A}}_\text{sym} := \frac{1}{2}\left(\widetilde{\mathbf{A}} + \widetilde{\mathbf{A}}^T\right).$$
Then we modify $\widetilde{\mathbf{A}}_\text{sym}$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
\label{eq:rational_matrix_function}
\widetilde{\mathbf{A}}_\text{sym}^+ := c_0 \mathbf{I} + c_1 \widetilde{\mathbf{A}}_\text{sym} + c_2 \left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\widetilde{\mathbf{A}}_\text{sym}$. The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the spectrum of $\widetilde{\mathbf{A}}_\text{sym}^+$ has certain desirable properties. Later in this section we will describe how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\widetilde{\mathbf{A}}_\text{sym}^+$. Forming $\widetilde{\mathbf{A}}_\text{sym}^+$ requires computing an H-matrix inverse, $\left(\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is fine to compute the inverse because $\widetilde{\mathbf{A}}_\text{sym} + \mu \mathbf{I}$ is an H-matrix. There exist fast, stable, scalable, and widely used algorithms for inverting H-matrices \cite{Hmatrixinverse}.

@@

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $f(0)=0$ is the unique minimum of $f$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $f$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $f(\widetilde{\mathbf{A}}_\text{sym}) = \widetilde{\mathbf{A}}_\text{sym}^+$ is positive semi-definite. The equation $f(0)=0$ implies that the null space of $\widetilde{\mathbf{A}}_\text{sym}^+$ contains the null-space of $\widetilde{\mathbf{A}}_\text{sym}$. This is important when the operator $\mathcal{A}$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. The equation $f(\lambda_\text{max})=\lambda_\text{max}$ forces the moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ to be close to the corresponding moderate and large positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$. I.e., $f$ modifies the ``important'' part of the spectrum of $\widetilde{\mathbf{A}}_\text{sym}$ as little as possible. The equation $f(\lambda_\text{min})=|\lambda_\text{min}|$ ensures that the negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $f$. There is a tradeoff: if we choose a larger value for $f(\lambda_\text{min})$, then the positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}^+$ will be closer to the corresponding positive eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ (good), but the erroneous negative eigenvalues of $\widetilde{\mathbf{A}}_\text{sym}$ will be amplified more in magnitude (bad). This is shown in Figure FIG. We find that $f(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.
